---
title: "Survival_ana"
format: html
editor: visual
---

## Survival Analysis

This notebook demonstrates survival analysis applied to data generated from a car racing simulation. In the simulation, cars drive along a track and are subject to random lightning strikes that may destroy them. The distance driven until destruction is treated as the survival time.

The analysis is structured in two parts.

-   In the first part, a Kaplan-Meier estimator and an exponential survival model are fitted to the data from a single group of cars.

-   In the second part, the simulation is extended to two teams with different survival probabilities upon a lightning strike. A Cox Proportional Hazards model is used to estimate and test the hazard ratio between the two teams.

Since the data is generated from a simulation with known parameters, the statistical results can be verified against the true underlying values, providing a useful sanity check of the methodology. You can find the html page used for data generation in the same folder.

```{r}
library(survival)
library(dplyr)
library(ggplot2)
library(ggsurvfit)
library(flexsurv)
```

## Preparing Data

We load the dataset and prepare the data:

```{r}
df <- read.table("race_results.csv", sep = ",", header = TRUE)
dim(df)
```

We have the following columns:

-   a car column to identify each car

-   a status column if a car was destroyed or not

-   a column counting the laps

-   a column with the meters driven by the car until the end of the simulation or destruction.

```{r}
head(df)
```

```{r}
# Clean column names
colnames(df) <- c("car", "status", "laps", "distance_m")
colnames(df)
```

```{r}
ggplot(df, aes(x = distance_m)) +
  geom_histogram(binwidth = 200, fill = "steelblue", color = "white") +
  labs(x = "Distance (m)", y = "Count", title = "Histogram of Distance") +
  theme_minimal()
```

```{r}
# Create binary event indicator: 1 = destroyed, 0 = censored (alive)
df <- df %>%
  mutate(
    event = ifelse(status == "destroyed", 1, 0),
    
    # Parse laps: strip the "+" suffix and convert to numeric
    laps_numeric = as.numeric(gsub("\\+", "", laps))
  )
```

```{r}
df
```

We use distance column as time in the model. Since all cars are driving with the same speed, we can do this. The event column indicates if a car is destroyed or not. A value of one indicates that a car was destroyed.

```{r}
# Create the Surv object (using distance as the time scale)
surv_obj <- Surv(time = df$distance_m, event = df$event)

# Quick sanity check
summary(surv_obj)
```

## Analysis

Let us start by looking at the Kaplan-Meier model. The curve in the plot below is an estimation of the Survival function $S(t)=1-F(t)$ . This model is non-parametric, i.e. no assumption about the distribution of the survival is made.

```{r}
# Fit the Kaplan-Meier model
km_fit <- survfit(surv_obj ~ 1, data = df)

# Summary with confidence intervals
summary(km_fit)

# Plot using ggsurvfit
survfit2(Surv(distance_m, event) ~ 1, data = df) %>%
  ggsurvfit(linewidth = 1) +
  add_confidence_interval() +
  add_risktable() +
  add_censor_mark() +
  scale_x_continuous(
    name = "Distance (m)",
    breaks = seq(0, 7000, by = 1000)
  ) +
  scale_y_continuous(
    name = "Survival Probability",
    limits = c(0, 1),
    labels = scales::percent
  ) +
  labs(
    title = "Kaplan-Meier Survival Curve",
    subtitle = "Survival probability of cars by distance driven"
  ) +
  theme_minimal(base_size = 13)
```

From the simulation we know that

-   lightning strikes seem to occur everywhere along the race with same probability

-   the rate of destruction remains constant, i.e. cars are not slowing down and the frequency of lightning strikes stays the same.

This suggests that the hazard function $h(t)$ is a constant, i.e. $h(t)=\lambda$ for some $\lambda>0$. The hazard function and the survival function $S(t)$ are related via:

$$
\begin{aligned}
S(t) &= \exp(-\int_0^t h(u)du), 
\end{aligned}
$$

Consequently, we have:

$$
\begin{aligned}
S(t) &= \exp(-\int_0^t \lambda du), \\
&= \exp(-t \lambda) 
\end{aligned}
$$

Since $F(t) = 1 - S(t)$ we have $f(t) =\lambda \exp(-t \lambda)$. I.e. the survival times follow an exponential distribution.

Therefore, we fit a model where the survival times follow an exponential distribution.

```{r}
# Fit the exponential model
# The dist parameter specifies the parametric distribution assumed for the survival times. 
# dist = "exp" means we assume the survival times follow an exponential distribution
# This corresponds to a constant hazard $h(t) = \lambda$.
exp_fit <- flexsurvreg(surv_obj ~ 1, 
                       data = df, 
                       dist = "exp")

# Print the model summary
print(exp_fit)
```

Five samples were censored corresponding to the five cars still driving when the simulation was stopped. **The rate estimate** is $\hat{\lambda} = 1.92 \times 10^{-4}$ per meter, with 95% CI $[1.16 \times 10^{-4},\ 3.18 \times 10^{-4}]$. This means that

$$
\begin{aligned}
\hat S(t) &= \exp(-0.000192 t), 
\end{aligned}
$$

We can read from this for example:

-   The model predicts that on average a car passes $1/\hat{\lambda} = 5208.333$ meters before destruction.

-   According to the model the probability that a car drives more than $20000$ meters is roughly $2$ percentage.

With only 15 events the confidence interval spans almost a factor of 3 ($1.16$ to $3.18$), so the estimate is quite uncertain.

We can plot the Kaplan-Meier plots and the exponential fit together to assess the goodness of the fit visually:

```{r}

# Extract lambda
lambda <- exp_fit$res["rate", "est"]
cat("Lambda (hazard rate):", lambda, "\n")
cat("Mean survival distance (1/lambda):", 1/lambda, "m\n")

# Plot KM curve with exponential model overlaid
plot(km_fit,
     xlab = "Distance (m)",
     ylab = "Survival Probability",
     main = "KM Curve vs Exponential Model",
     conf.int = TRUE,
     col = "steelblue",
     lwd = 2)

# Overlay the fitted exponential curve
t_seq <- seq(0, max(df$distance_m), length.out = 500)
lines(t_seq, exp(-lambda * t_seq), 
      col = "firebrick", lwd = 2, lty = 2)

legend("topright",
       legend = c("Kaplan-Meier", "Exponential fit"),
       col = c("steelblue", "firebrick"),
       lwd = 2, lty = c(1, 2))
```

As expected we see that the exponential model seems to capture the data quite well.

## Covariates

Let us look at a similar dataset but with covariates. Now, there are two teams. A car either belongs to team 1 or to team 2.

```{r}
df <- read.table("race_results_teams.csv", sep = ",", header = TRUE)
colnames(df) <- c("Car", "Team", "Status", "Laps", "distance_m")
df <- df %>%
  mutate(
    event = ifelse(Status == "destroyed", 1, 0),
    Team = ifelse(Team == "Team 1", 1, 0),
    # Parse laps: strip the "+" suffix and convert to numeric
    laps_numeric = as.numeric(gsub("\\+", "", Laps))
  )
dim(df)
```

```{r}
df
```

Again we look at the Kaplan-Meier curves, but this time stratified over the two teams.

```{r}
# Stratified Kaplan-Meier by team
survfit2(Surv(distance_m, event) ~ Team, data = df) %>%
  ggsurvfit(linewidth = 1) +
  add_confidence_interval() +
  add_risktable() +
  add_censor_mark() +
  scale_x_continuous(
    name = "Distance (m)",
    breaks = seq(0, 7000, by = 1000)
  ) +
  scale_y_continuous(
    name = "Survival Probability",
    limits = c(0, 1),
    labels = scales::percent
  ) +
  labs(
    title = "Kaplan-Meier by Team",
    subtitle = "Team 1 = 1, Team 2 = 0"
  ) +
  theme_minimal(base_size = 13)

```

This time, we will apply a Cox Proportional Hazard model which is semi-parametric. In this model we assume the hazard function takes the form:

$$
\begin{aligned}
h_i(t) &= h_0(t) \exp(\beta x_i), 
\end{aligned}
$$

The function $h_0$ is not known and $x_i$ represents the covariates for a sample. If we look at a sample of the first team and second team, let us say $i$ and $j$ , we can compute the ratio of their hazards even though we do not know $h_0$. That is:

$$
\begin{aligned}
\frac{h_i(t)}{h_j(t)} &= \exp(\beta x_i - \beta x_j), 
\end{aligned}
$$

Since Team 1 = 1 and Team 2 = 0, when comparing team 1 ($x_i = 1$) to team 2 ($x_j = 0$) the hazard ratio simplifies to $e^{\beta}$.

Note, that this does not mean that ratio of the survival functions is $e^{\beta}$. The hazard function is related to the survival function of the model via:

$$
\begin{aligned}
S(t) &= \exp(-\int_0^t h(u)\,du),
\end{aligned}
$$

and therefore

$$
\begin{aligned}
S_i(t) &= \exp(-\int_0^t h_0(u) \exp(\beta x_i)du), \\
 &= \exp(-\exp(\beta x_i) \int_0^t h_0(u) du), \\
 &= \left[\exp(-H_0(t))\right]^{\exp(\beta x_i)}, \\
 &= S_0(t)^{\exp(\beta x_i)}, \\
\end{aligned}
$$

Consequently, we have for the two groups;

$$
\begin{aligned}
S_1(t) &= S_0(t)^{\exp(\beta)}, \\
S_2(t) &= S_0(t),
\end{aligned}
$$

So the ratio of the survival functions of the two groups depends on $t$ and is therefore not constant over time. Instead we have

$$
\begin{aligned}
S_1(t) &= S_2(t)^{\exp(\beta)}, 
\end{aligned}
$$

Let us now fit the model:

```{r}
# Fit the Cox Proportional Hazards model
cox_null <- coxph(Surv(distance_m, event) ~ 1, data = df)
cox_team <- coxph(Surv(distance_m, event) ~ Team, data = df)
anova(cox_null, cox_team)
```

The p-value is 0.001 indicating that it is justified to include the categorical variable "Team" into the model.

```{r}
summary(cox_team)
```

We can see for example that the hazard ratio between the teams is $0.1980$. Since the confidence interval does not include $1$ the difference is significantly different under a level of $0.05$ .

Using the above calculations showing the relation between the survival curves in the two groups we can plot the survival curve of the second group against the survival curve of the first group.

```{r}
s2_seq <- seq(0, 1, length.out = 200)
theory  <- data.frame(S2 = s2_seq, S1 = s2_seq^0.198)

ggplot(theory, aes(x = S2, y = S1)) +
  geom_line(color = "firebrick") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey50") +
  labs(x = expression(S[2](t)), 
       y = expression(S[1](t)),
       title = "Survival-Survival Plot",
       subtitle = expression(S[1] == S[2]^{0.198})) +
  theme_minimal()


```

The proportional hazard model assumes that the hazard ratio between the two groups is constant over time. We can test this assumption:

```{r}
# Test the proportional hazards assumption via Schoenfeld residuals
ph_test <- cox.zph(cox_team)
plot(ph_test, main = "Schoenfeld Residuals — Team")
```

```{r}
# p > 0.05 means the PH assumption holds
print(ph_test)
```

Since $p > 0.05$ the PH assumption is not rejected.

In our case, since we simulated our data, we know actually what the hazard ratio is. The hazard function $h(t)$ is the rate of destruction given a car has made it until $t$. I.e. of the cars that have survived up to distance $t$, the expected fraction to be destroyed in the next small stretch $dt$ is $h(t)$. When creating the simulated data we have a parameter determining the probability that a car survives a lightning strike. We claim for Team 2 this parameter was zero and for Team 1 was $p=0.72$. That is a car in Team 1 is destroyed with $1-p$ . So we would expect that the hazard ratio $\frac{h_1(t)}{h_2(t)}$ between the two teams is $1-p$ and consequently $\beta = \log(1-p) = -1.272966$. We can test this hypothesis:

```{r}
# Fit model with beta fixed at log(0.28) using offset
df$offset_val <- log(1 - 0.72) * df$Team
# The offset in a Cox model allows you to fix a coefficient at a specific value rather than estimating it from the data.
cox_fixed <- coxph(Surv(distance_m, event) ~ offset(offset_val), data = df)
cox_free  <- coxph(Surv(distance_m, event) ~ Team, data = df)

anova(cox_fixed, cox_free)
```

As expected, there is no statistically significant evidence at the $\alpha = 0.05$ level that the estimated hazard ratio differs from the true value $1 - p = 0.28$.

