---
title: "Survival_ana"
format: html
editor: visual
---

## Survival Analysis

Data is taken from the simulation in the same folder.

```{r}
library(survival)
library(dplyr)
library(ggplot2)
library(ggsurvfit)
library(flexsurv)
```

## Preparing Data

We load the dataset and prepare the data:

```{r}
df <- read.table("race_results.csv", sep = ",", header = TRUE)
dim(df)
```

You can add options to executable code like this

```{r}
head(df)
```

The `echo: false` option disab

```{r}
# Clean column names
colnames(df) <- c("car", "status", "laps", "distance_m")
colnames(df)
```

les the printing of code (only output is displayed).

```{r}
df$distance_m
```

```{r}
# Create binary event indicator: 1 = destroyed, 0 = censored (alive)
df <- df %>%
  mutate(
    event = ifelse(status == "destroyed", 1, 0),
    
    # Parse laps: strip the "+" suffix and convert to numeric
    laps_numeric = as.numeric(gsub("\\+", "", laps))
  )
```

```{r}
df
```

```{r}

# Create the Surv object (using distance as the time scale)
surv_obj <- Surv(time = df$distance_m, event = df$event)

# Quick sanity check
summary(surv_obj)
```

## Analysis

Let us start by looking at the Kaplan-Meier model. The curve in the plot below is an estimation of the Survival function $S(t)=1-F(t)$ . This model is non-parametric, i.e. no assumption about the distribution of the survival is made.

```{r}
# Fit the Kaplan-Meier model
km_fit <- survfit(surv_obj ~ 1, data = df)

# Summary with confidence intervals
summary(km_fit)

# Plot using ggsurvfit
survfit2(Surv(distance_m, event) ~ 1, data = df) %>%
  ggsurvfit(linewidth = 1) +
  add_confidence_interval() +
  add_risktable() +
  add_censor_mark() +
  scale_x_continuous(
    name = "Distance (m)",
    breaks = seq(0, 7000, by = 1000)
  ) +
  scale_y_continuous(
    name = "Survival Probability",
    limits = c(0, 1),
    labels = scales::percent
  ) +
  labs(
    title = "Kaplan-Meier Survival Curve",
    subtitle = "Survival probability of cars by distance driven"
  ) +
  theme_minimal(base_size = 13)
```

We observe that the Survival curve looks fairly linear. This suggests that:

-   lightning strikes seem to occur everywhere along the race with same probability

-   the rate of destruction remains constant

```{r}
# Fit the exponential model
exp_fit <- flexsurvreg(Surv(distance_m, event) ~ 1, 
                       data = df, 
                       dist = "exp")

# Print the model summary
print(exp_fit)
```

```{r}

# Extract lambda
lambda <- exp_fit$res["rate", "est"]
cat("Lambda (hazard rate):", lambda, "\n")
cat("Mean survival distance (1/lambda):", 1/lambda, "m\n")

# Plot KM curve with exponential model overlaid
plot(km_fit,
     xlab = "Distance (m)",
     ylab = "Survival Probability",
     main = "KM Curve vs Exponential Model",
     conf.int = TRUE,
     col = "steelblue",
     lwd = 2)

# Overlay the fitted exponential curve
t_seq <- seq(0, max(df$distance_m), length.out = 500)
lines(t_seq, exp(-lambda * t_seq), 
      col = "firebrick", lwd = 2, lty = 2)

legend("topright",
       legend = c("Kaplan-Meier", "Exponential fit"),
       col = c("steelblue", "firebrick"),
       lwd = 2, lty = c(1, 2))
```

```{r}
```

## Covariates

Let us look at a similar dataset but with covariates. Now, there are two teams. A car either belongs to team 1 or to team 2.

```{r}
df <- read.table("race_results_teams.csv", sep = ",", header = TRUE)
colnames(df) <- c("Car", "Team", "Status", "Laps", "distance_m")
df <- df %>%
  mutate(
    event = ifelse(Status == "destroyed", 1, 0),
    Team = ifelse(Team == "Team 1", 1, 0),
    # Parse laps: strip the "+" suffix and convert to numeric
    laps_numeric = as.numeric(gsub("\\+", "", Laps))
  )
dim(df)
```

```{r}
df
```

Again we look at the Kaplan-Meier curves, but this time stratified over the two teams.

```{r}
# Stratified Kaplan-Meier by team
survfit2(Surv(distance_m, event) ~ Team, data = df) %>%
  ggsurvfit(linewidth = 1) +
  add_confidence_interval() +
  add_risktable() +
  add_censor_mark() +
  scale_x_continuous(
    name = "Distance (m)",
    breaks = seq(0, 7000, by = 1000)
  ) +
  scale_y_continuous(
    name = "Survival Probability",
    limits = c(0, 1),
    labels = scales::percent
  ) +
  labs(
    title = "Kaplan-Meier by Team",
    subtitle = "Team 1 = 1, Team 2 = 0"
  ) +
  theme_minimal(base_size = 13)

```

We will apply a Cox Proportional Hazard model which is semi-parametric. In this model wee assume the hazard function takes the form:

$$
\begin{aligned}
h_i(t) &= h_0(t) \exp(\beta x_i), 
\end{aligned}
$$

The function $h_0$ is not known and $x_i$ represents the covariates for a sample. If we look at a sample of the first team and second team, let us say $i$ and $j$ , we can compute the ratio of their hazards even though we do not know $h_0$. That is:

$$
\begin{aligned}
\frac{h_i(t)}{h_j(t)} &= \exp(\beta x_i - \beta x_j), 
\end{aligned}
$$

Since Team 1 = 1 and Team 2 = 0, when comparing team 1 ($x_i = 1$) to team 2 ($x_j = 0$) the hazard ratio simplifies to $e^{\beta}$.

Let us fit the model:

```{r}
# Fit the Cox Proportional Hazards model
cox_null <- coxph(Surv(distance_m, event) ~ 1, data = df)
cox_team <- coxph(Surv(distance_m, event) ~ Team, data = df)
summary(cox_team)
```

We can see for example that the hazard ratio between the teams is $0.1980$. Since the confidence interval does not include $1$ the difference is signficantley different under a level of $0.05$ .

The proportional hazard model assumes that the hazard ratio between the two groups is constant over time.

```{r}
# Test the proportional hazards assumption via Schoenfeld residuals
ph_test <- cox.zph(cox_team)
plot(ph_test, main = "Schoenfeld Residuals — Team")
```

```{r}
# p > 0.05 means the PH assumption holds
print(ph_test)
```

Since $p > 0.05$ the PH assumption is not rejected.

```{}
```
